A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-05-11 paper/paper-jsdse/paper-jsdse.tex
section 3 done

Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003). We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the \emph{purposes of studying racism, not race} (Fullilove 1998). There are a number of data sets provided by Singer and Willett (2003), and we focus only on this one because it has captivated our attention for a number of years. We use it in our own teaching of longitudinal data analysis and would very much like to use data since 1994.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The wages data of the whole NLSY79 cohort, including females.
\item
  A separate table of the demographic data of the whole NLSY79 cohort.
\item
  The subset of wages data in (1), which is the wages of high school dropouts' as a refreshed version of Singer and Willett (2003)'s data.
\end{enumerate}

@@

In the context of official statistics, M. van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables). What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain. In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida} and Section \ref{censor}.

@@

Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the \texttt{tidyverse} suite of packages (Wickham et al. 2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham et al. 2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{https://github.com/numbats/yowie/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

Further, the object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}). The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981. We checked for consistency between the two versions and found no discrepancy where the responses were recorded in both 1979 and 1981. The age was then calculated using the birth year.

The next step is processing the highest grade completed. There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis. The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively. For each individual, there is only one value of \texttt{hgc} and \texttt{hgc\_i}. This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on). We decide to include both the factor and integer values to be used as examples in various classroom demonstrations, such as to explain different data types. Besides, it can also be used as the demonstration of different type of data visualization which is determined by the data type. It can also be used for teaching the modeling of cross sectional data.

@@

The hourly wages are in the variables beginning with \texttt{HRP} in \texttt{new\_data\_qnames}. As a respondent may have multiple jobs, the \texttt{mean\_hourly\_wage} is computed as a weighted average of the hourly wage for each job with the number of hours worked for each job as weights (provided that the information on the number of hours is available); if the number of hours worked for any job is missing, then the \texttt{mean\_hourly\_wage} is computed as a simple average of all available hourly wages. Prior to computing the mean hourly wage, we undertook a number of steps to treat unusual observations as described below:

\begin{itemize}
\tightlist
\item
  If the hourly rate is recorded as 0, we set wage as missing.
\item
  If the total hours of worked for the corresponding job is greater than 84 hours, we set the wage and hours worked as missing.
\end{itemize}

@@

There are many steps and decisions made to go from raw to input to valid data. Figure \ref{fig:flow-chart-blind} summarizes these in order to create a refreshed wages data set.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \texttt{id}: A unique individual's ID number. This is the \texttt{key} of the data as we saved the data as a \texttt{tsibble} object.
\item
  \texttt{year}: The year the observation was taken. This is the \texttt{index} of the data.
\item
  \texttt{wage}: The mean of the hourly wages the individual gets at each of their different jobs. The value could be a weighted or an arithmetic mean. The weighted mean is used when the information of hours of work as the weight is available. The mean hourly wage could also be a predicted value if the original value is considered influential by the robust linear regression as part of data cleaning.
\item
  \texttt{age\_1979}: The age of the individual in 1979.
\item
  \texttt{sex}: Sex of the individual, f = Female and m = Male.
\item
  \texttt{race}: Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
\item
  \texttt{grade}: Integer value of the highest grade completed corresponding to the survey year.
\item
  \texttt{hgc}: The highest grade completed ever.
\item
  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
\item
  \texttt{hgc\_1979}: The highest grade completed in 1979 (integer value).
\item
  \texttt{ged}: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
\item
  \texttt{njobs}: Number of jobs that an individual has.
\item
  \texttt{hours:}: The total number of hours the individual usually works per week.
\item
  \texttt{stwork}: The year when the individual started to work.
\item
  \texttt{yr\_wforce}: The length of time in the workforce in years (\texttt{year} - \texttt{stwork}).
\item
  \texttt{exp}: Work experience, i.e., the number of years of working.
\item
  \texttt{is\_wm}: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
\item
  \texttt{is\_pred}: Whether the mean hourly wage is a predicted value of RLM or not.
\end{enumerate}

@@

\begin{itemize}
\item
  \textbf{Codes}: R script to reproduce data tidying and cleaning is available at \url{https://github.com/numbats/yowie/articles/process-data.html}. The code for the extreme value handling with robust linear fits is in \url{https://github.com/numbats/yowie/articles/input-to-valid-data.html}.
\item
  \textbf{R Package}:\texttt{yowie} is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort. This package can be accessed from \url{https://github.com/numbats/yowie}.
\item
  \textbf{shiny app}: An interactive \texttt{shiny} web app to visualise the effect of selecting different weight threshold for substituting the wages data to its predicted value from a fit of the robust linear regression model. This app can be accessed at \url{https://ebsmonash.shinyapps.io/yowie_app/} with the source code provided \url{https://github.com/numbats/yowie/tree/master/app}.
\end{itemize}

>>>>

Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003). We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the \emph{purposes of studying racism, not race} (Fullilove 1998). There are a number of data sets provided by Singer and Willett (2003), but we focus only on wages data because it has captivated our attention for a number of years in our own teaching of longitudinal data analysis.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The wages data of the whole NLSY79 cohort, including females,
\item
  A separate table of the demographic data of the whole NLSY79 cohort, and
\item
  The subset of wages data in (1), which is the wages of high school dropouts' as a refreshed version of Singer and Willett (2003)'s data.
\end{enumerate}

@@

In the context of official statistics, M. van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables). What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain. In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida} and Section \ref{censor}.

@@

Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the \texttt{tidyverse} suite of packages (Wickham et al. 2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham et al. 2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{https://numbats.github.io/yowie/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

Furthermore, the object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}). The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981. We checked for consistency between the two versions and found no discrepancy where the responses were recorded in both 1979 and 1981. The age was then calculated using the birth year.

The next step is processing the highest grade completed. There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis. The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, encoded as a factor and integer variable, respectively. For each individual, there is only one value of \texttt{hgc} and \texttt{hgc\_i}. This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on). We decide to include both the factor and integer variables to be used as examples in various classroom demonstrations to explain different data types say. Besides this, it can also serve as a demonstration of different type of data visualization ascribed by the data type and for teaching the modeling of cross sectional data.

@@

The hourly wages are in the variables beginning with \texttt{HRP} in \texttt{new\_data\_qnames}. As a respondent may have multiple jobs, the \texttt{mean\_hourly\_wage} is computed as a weighted average of the hourly wage for each job with the number of hours worked for each job as weights (provided that the information on the number of hours is available); if the number of hours worked for any job is missing, then the \texttt{mean\_hourly\_wage} is computed as a simple average of all available hourly wages. Prior to computing the mean hourly wage, we undertook a number of steps to treat extreme observations as described below:

\begin{itemize}
\tightlist
\item
  If the hourly rate is recorded as 0, we set wage as missing, and
\item
  If the total hours of worked per week for the corresponding job is greater than 84 hours, we set the wage and hours worked as missing.
\end{itemize}

@@

There are many steps and decisions made to go from raw to input to valid data. Figure \ref{fig:flow-chart} summarizes these in order to create a refreshed wages data set.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \texttt{id}: A unique individual's ID number. This is the \texttt{key} of the data as we saved the data as a \texttt{tsibble} object.
\item
  \texttt{year}: The year the observation was taken. This is the \texttt{index} of the data.
\item
  \texttt{wage}: The mean of the hourly wages the individual gets at each of their different jobs. The value could be a weighted or an arithmetic mean. The weighted mean is used when the information of hours of work as the weight is available. The mean hourly wage could also be a predicted value if the original value is considered influential by the robust linear regression as part of data cleaning.
\item
  \texttt{age\_1979}: The age of the individual in 1979.
\item
  \texttt{sex}: Sex of the individual, f = Female and m = Male.
\item
  \texttt{race}: Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
\item
  \texttt{grade}: Integer value of the highest grade completed corresponding to the survey year.
\item
  \texttt{hgc}: The highest grade completed ever.
\item
  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
\item
  \texttt{hgc\_1979}: The highest grade completed in 1979 (integer value).
\item
  \texttt{ged}: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
\item
  \texttt{njobs}: Number of jobs that an individual has.
\item
  \texttt{hours}: The total number of hours the individual usually works per week.
\item
  \texttt{stwork}: The year when the individual started to work.
\item
  \texttt{yr\_wforce}: The length of time in the workforce in years (\texttt{year} - \texttt{stwork}).
\item
  \texttt{exp}: Work experience, i.e., the number of years of working.
\item
  \texttt{is\_wm}: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
\item
  \texttt{is\_pred}: Whether the mean hourly wage is a predicted value of RLM or not.
\end{enumerate}

@@

\begin{itemize}
\item
  \textbf{Codes}: R script to reproduce data tidying and cleaning is available at \url{https://numbats.github.io/yowie/articles/process-data.html}. The code for the extreme value handling with robust linear fits is in \url{https://numbats.github.io/yowie/articles/input-to-valid-data.html}.
\item
  \textbf{R Package}:\texttt{yowie} is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort. This package can be accessed from \url{https://github.com/numbats/yowie}.
\item
  \textbf{shiny app}: An interactive \texttt{shiny} web app to visualise the effect of selecting different weight threshold for substituting the wages data to its predicted value from a fit of the robust linear regression model. This app can be accessed at \url{https://ebsmonash.shinyapps.io/yowie_app/} with the source code provided \url{https://github.com/numbats/yowie/tree/master/app}.
\end{itemize}

### 2022-05-11 paper/paper.Rmd
section 3 done

```{r censor, include = FALSE}
CENSOR <- FALSE
yowie <- ifelse(CENSOR, "\\texttt{[CENSORED]}", "\\texttt{yowie}")
yowie_repo <- ifelse(CENSOR, "CENSORED", "https://github.com/numbats/yowie")
yowie_pkgdown <- ifelse(CENSOR, "CENSORED", "https://github.com/numbats/yowie")
yowie_code <- ifelse(CENSOR, "CENSORED", "https://github.com/numbats/yowie-paper/data-raw")
yowie_shiny <- ifelse(CENSOR, "CENSORED", "https://ebsmonash.shinyapps.io/yowie_app/")
censor_word <- function(word) ifelse(CENSOR, "CENSORED", word)
```

@@

Further, the object `new_data_qnames`, contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981`, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`). The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981. We checked for consistency between the two versions and found no discrepancy where the responses were recorded in both 1979 and 1981. The age was then calculated using the birth year.

@@

The next step is processing the highest grade completed. There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis. The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as `hgc` and `hgc_i`, for the factor and integer type, respectively. For each individual, there is only one value of `hgc` and `hgc_i`. This variable is obtained from `new_data_qnames` with the name `HGC_EVER_XRND` and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as `hgc_i` and recoded as a factor to give `hgc` (e.g., 10th grade, 11th grade, and so on). We decide to include both the factor and integer values to be used as examples in various classroom demonstrations, such as to explain different data types. Besides, it can also be used as the demonstration of different type of data visualization which is determined by the data type. It can also be used for teaching the modeling of cross sectional data.

@@

The hourly wages are in the variables beginning with `HRP` in `new_data_qnames`. As a respondent may have multiple jobs, the `mean_hourly_wage` is computed as a weighted average of the hourly wage for each job with the number of hours worked for each job as weights (provided that the information on the number of hours is available); if the number of hours worked for any job is missing, then the `mean_hourly_wage` is computed as a simple average of all available hourly wages. Prior to computing the mean hourly wage, we undertook a number of steps to treat unusual observations as described below:

* If the hourly rate is recorded as 0, we set wage as missing.
* If the total hours of worked for the corresponding job is greater than 84 hours, we set the wage and hours worked as missing.

@@

  11. `ged`: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
  12. `njobs`: Number of jobs that an individual has.
  13. `hours:`: The total number of hours the individual usually works per week.
  14. `stwork`: The year when the individual started to work.
  15. `yr_wforce`: The length of time in the workforce in years (`year` - `stwork`).
  16. `exp`: Work experience, i.e., the number of years of working. 
  17. `is_wm`: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
  18. `is_pred`: Whether the mean hourly wage is a predicted value of RLM or not.

>>>>

```{r censor, include = FALSE}
CENSOR <- FALSE
yowie <- ifelse(CENSOR, "\\texttt{[CENSORED]}", "\\texttt{yowie}")
yowie_repo <- ifelse(CENSOR, "CENSORED", "https://github.com/numbats/yowie")
yowie_pkgdown <- ifelse(CENSOR, "CENSORED", "https://numbats.github.io/yowie")
yowie_code <- ifelse(CENSOR, "CENSORED", "https://github.com/numbats/yowie-paper/data-raw")
yowie_shiny <- ifelse(CENSOR, "CENSORED", "https://ebsmonash.shinyapps.io/yowie_app/")
censor_word <- function(word) ifelse(CENSOR, "CENSORED", word)
```

@@

Furthermore, the object `new_data_qnames`, contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981`, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`). The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981. We checked for consistency between the two versions and found no discrepancy where the responses were recorded in both 1979 and 1981. The age was then calculated using the birth year.

@@

The next step is processing the highest grade completed. There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis. The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as `hgc` and `hgc_i`, encoded as a factor and integer variable, respectively. For each individual, there is only one value of `hgc` and `hgc_i`. This variable is obtained from `new_data_qnames` with the name `HGC_EVER_XRND` and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as `hgc_i` and recoded as a factor to give `hgc` (e.g., 10th grade, 11th grade, and so on). We decide to include both the factor and integer variables to be used as examples in various classroom demonstrations to explain different data types say. Besides this, it can also serve as a demonstration of different type of data visualization ascribed by the data type and for teaching the modeling of cross sectional data.

@@

The hourly wages are in the variables beginning with `HRP` in `new_data_qnames`. As a respondent may have multiple jobs, the `mean_hourly_wage` is computed as a weighted average of the hourly wage for each job with the number of hours worked for each job as weights (provided that the information on the number of hours is available); if the number of hours worked for any job is missing, then the `mean_hourly_wage` is computed as a simple average of all available hourly wages. Prior to computing the mean hourly wage, we undertook a number of steps to treat extreme observations as described below:

* If the hourly rate is recorded as 0, we set wage as missing, and
* If the total hours of worked per week for the corresponding job is greater than 84 hours, we set the wage and hours worked as missing.

@@

  11. `ged`: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
  12. `njobs`: Number of jobs that an individual has.
  13. `hours`: The total number of hours the individual usually works per week.
  14. `stwork`: The year when the individual started to work.
  15. `yr_wforce`: The length of time in the workforce in years (`year` - `stwork`).
  16. `exp`: Work experience, i.e., the number of years of working. 
  17. `is_wm`: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
  18. `is_pred`: Whether the mean hourly wage is a predicted value of RLM or not.
